Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science

### 2025-09-16 conference-materials/2025_posit-conf/draft.qmd
Update conference presentation slides with refined content and styling
- Streamlined speaker notes and slide text for better flow
- Updated dashboard metrics visualization with cleaner 4-panel layout
- Improved responsive styling for embedded GIFs and column layouts
- Simplified code examples and removed unnecessary package dependencies
- Enhanced readability of tables and positioning of overlay elements

---
title: "Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science"
author: "Jeremy Winget, PhD"
date: "2025-09-17"
date-format: "MMMM D, YYYY"
execute:
  echo: true
  eval: true
format:
  revealjs:
    code-block-background: true
    highlight-style: github
    slide-number: false
    transition: fade
    auto-animate: true
    width: 1600
    height: 900
    footer: "posit::conf(2025) | jrwinget\\.com"
    title-slide-attributes:
      data-background-image: "img/posit-conf.png"
      data-background-position: "top left"
      data-background-repeat: "no-repeat"
      data-background-color: "#213d4f"
      data-background-opacity: "0.25"
      data-background-size: "45%"
      data-notes: |
        - [10 seconds; smile warmly at audience]
        - Good morning everyone! Let's jump right in...
---

@@

:::: {.columns .center}
::: {.column width="45%" .fragment .center}
<div class="tenor-gif-embed" data-postid="596341483704119392" data-share-method="host" data-aspect-ratio="1.32447" data-width="700px"><a href="https://tenor.com/view/monkey-computer-work-bug-developer-gif-596341483704119392">Monkey Computer GIF</a>from <a href="https://tenor.com/search/monkey-gifs">Monkey GIFs</a></div><script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::

::: {.column width="45%" .fragment .center}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.32447" data-width="925px"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div><script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::
::::

::: {.notes}
- [30 seconds]
- Who here has opened a dashboard with 18 filters across 6 tabs...and wondered where to even start?...Or maybe what they were even looking at?
- I see some knowing looks
- Now I'd like you to imagine you're NOT a data scientist...
- You're an executive who needs to make a decision in the next 10 mins
- [CLICK] This might be you...frantically clicking, hoping something makes sense
- [CLICK] Or maybe you have some...bigger feelings
- The truth is...we've all probably felt this way towards a dashboard...and it's never a good feeling
- [end by 40 secs]
:::

@@

::: {.notes}
- [60 seconds]
- So...welcome to my dashboard support group!...Let's do a quick check-in
- Raise your hand if you've designed a UI with 10+ filters
    - Keep it up if you've gone over 20?
- How about this...who's had a user ask for "just one more dropdown" when you KNEW it was just too much?
- [CLICK] If you're nodding along...you're among friends...No judgment here
- [CLICK] I'm Jeremy...and I've been there too
- I'm a psychologist by training...but probably not the kind you're thinking of
- See...a lot of folks think psychology is all about couches and therapy...but the discipline is powered by data
- [CLICK] My journey took me from studying how groups make decisions under pressure...to building the systems that caused that pressure...to finally figuring out how to fix them
- And after years of building dashboards that were technically great...but didn't seem to land with users...I had this realization
- [end by 1:40 mins]
:::

@@

::: {.notes}
- 20 seconds
- A lot of dashboard failures aren't technical...They're psychological
- They don't fail bc of the wrong chart type or database
- They fail because humans need to use them...and humans are complicated
- But we can engineer for psychology just as precisely as we engineer for performance
- Let me show you what I mean
- [end by 2:00 mins]
:::

## {background-image="img/ui-before1.png" background-size="contain" background-color="#2c3e50"}

::: {.text-box-overlay-dark .absolute left="-200" bottom="100"}
### Real client. Real dashboard. Real problems.

- 18 filters across 6 tabs
- Every possible data view
- Users' curiosity quickly vanished
    - And so did their visits

@@

::: {.notes}
- [45 seconds]
- This is a dashboard from an actual client...details blinded of course, but still accurate
- They had built a really impressive dashboard...
- Lots of filters...customized views...and some really cool visualizations for deep dives
- The client loved this bc..."users can analyze anything they want!"
- Except...they never did
- The opened the app...but didn't stay long before they completely deserted it
- They were finding creative workarounds...like Excel
- [CLICK] So we decided to roll out a minor update and collect anonymous usage data with telemetry for 30 days to get a better idea of what users were actually doing
- [end by 2:45]
:::

@@

::: {.notes}
- [35 seconds]
- And in this case...it painted a user journey like this cliff of despair
- 95% of users visited the URL...fantastic! Shows people are interested
- 88% opened the dashboard...also good...they're getting past the landing page
- But here's where it gets brutal...only 52% applied even one filter. Just one!
- And only 20% returned within 7 days
- This chart is behavioral science screaming "STOP DOING THIS!"
- But what needs fixing...and how do we do it?
- [end by 3:20]
:::

@@

::: {.notes}
- [75 seconds]
- With behavioral science!...which is the study of how humans think, feel and act...often in the context of decision making
- The field itself is a blend of psychology, economics, neuroscience, and others...
- But across all those areas we find people are often irrational...but in consistent ways
- For us as developers...this is actually great news!
- It means user behavior is predictable...and we can design for it
- [CLICK] Let me share a study that demonstrates our death by dropdown problem perfectly
- [CLICK] In this study...two researchers set up a sampling booth at a grocery store...and they varied the number of jams they on display
- [CLICK] In the big display condition...there was were 24 jams on the table...and in the small display condition...there were 6 jams
- [CLICK] The big display with 24 jams attracted more shoppers than the small display with 6 jams
- BUT...only 3% of folks who stopped at the larger display bought jam...whereas 30% of people bought jam in the small display condition
- [CLICK] That's a 10x conversion difference!
- So more choice attracted attention but paralyzed action...and filter-heavy dashboards do the exact same thing
- [end by 4:30]
:::

@@

- **Cognitive Load**
    - Too many choices overwhelm users' mental capacity
- **Progressive Disclosure**
    - Reveal complexity gradually as users need it
- **Framing Effects**
    - Same data can tell completely different stories
- **Choice Architecture**
    - Guide decisions without restricting options
:::

@@

- **Cognitive Load**
    - Start with 3 key filters, hide 15 advanced ones
- **Progressive Disclosure**
    - Show summary → detailed dive → comparisons
- **Framing Effects**
    - "Revenue up 15%" vs "Revenue missed by 5%"
- **Choice Architecture**
    - Smart defaults, logical grouping, visual hierarchy
:::
::::

::: {.notes}
- [75 seconds]
- Now...the paradox of choice is just one of many behavioral science concepts we can apply to dashboards
- But these aren't abstract theories...they're actually engineering principles we can leverage
- Cognitive Load...this means too many choices overwhelm users' mental capacity
    - Like RAM...your users' brains have processing limits...every filter costs processing power
- Progressive Disclosure...this means revealing complexity gradually as users need it
    - Show vital information first...and reveal more as needed
    - Like lazy loading for human cognition
- Framing Effects...this is how the same data can tell completely different stories
    - "Revenue up 15%" versus "Missed target by 5%"
    - Identical data...but the Proj Manager might see success while the CFO sees failure
- Choice Architecture...this means guiding the user without restricting them
    - Hide rarely used filters, use smart defaults
    - Like good API design...guide users down the happy path
- [CLICK] In practice...this means starting with 3 filters not 18...showing summaries before details...and using smart defaults
- This is how we engineer the user experience as carefully as we would the backend
- But what if you don't know behavioral science?
- [end by 5:35]
:::

@@

::: {.notes}
- [45 seconds]
- Well...I bet you do know something about road trips
- And engineering apps with behavioral science is a lot like planning a road trip
- You need to know your destination...anticipate the hazards...and make sure everyone gets there in one piece
- That's where the Behavioral Insight Design (BID) framework comes in...[CLICK]
    - It's like having GPS for your behavioral science journey
- Five simple stages that guide you through the process...and help prevent those dashboard disasters we've all seen
- What makes BID different is it's not some abstract design theory
- It's a concrete...step-by-step framework built on decades of research
- Each stage feeds into the next...creating this continuous feedback loop that keeps improving your dashboard
- Plus...the whole thing is already bundled into an R package that handles the heavy lifting
- [end by 6:20]
:::

@@

::: {.notes}
- [45 seconds]
- Meet bidux...designed specifically for our workflows as Shiny developers
- It works with any dashboard you've already built...no refactoring, no starting over
- Got telemetry? Great, it'll analyze actual user behavior and give you data-driven insights
- Or...if you don't have telemetry data...you can use simple verbal descriptions during the process
- Whatever inputs you provide...it will use them to auto-suggest improvements based on real research
- [CLICK] Think of it as having a behavioral scientist right in your console...always available, never judges, and doesn't charge by the hour
- Now...let me show you what this process actually looks like
- [end by 7:05]
:::

@@

::: {.notes}
- [65 seconds]
- First...we start with the INTERPRET stage...the why behind the dashboard
- What question does this dashboard answer?
- [CLICK] Because the central question drives the focus for your users
- If you can't state it in one simple sentence...they won't find it in a bunch of dropdowns
- [CLICK] And the data story helps create narrative structure for this
- Humans are hardwired for stories...not pivot tables
- Hook, context, tension, resolution...is the same structure folks are looking for with their key insights...so we build that in from the start
- [CLICK] And finally...the user personas help us devs take the user perspective
- The Product Manager isn't you...they don't think in SQL queries...They think in business outcomes
- So the dashboard needs to bridge that gap
- [CLICK] And when we put it all of that information together...the function creates a summary for you to review before moving on
- If it tells you anything is missing...it could also be a good time to check in with your users
- [end by 8:10]
:::

@@

::: {.notes}
- [75 seconds]
- Bc when we don't consider that context...we end up building critical friction points into our dashboards
- Going back to that opening example...here's what filter usage looked like
- 14 filters basically unused...4 virtually ignored
- This is technical debt visualized
- Each unused filter is maintenance overhead...and there are probably database queries running for nothing
- That's why the NOTICE stage focuses on finding these friction points systematically
- [CLICK] The package auto-suggests relevant behavioral theories with confidence scores
- Here...90% confidence it's Cognitive Load Theory...too many choices paralyzing users
- But it might suggest Choice Overload...Analysis Paralysis...or Decision Fatigue
- Each theory comes with specific, implementable solutions
- This transforms "it's confusing" into specific engineering problems...like "cognitive load exceeds working memory"
- The package also links each theory to specific Shiny components that address it
- So you're not guessing anymore...you're addressing identified psychological barriers with targeted code changes
- [end by 9:25]
:::

@@

::: {.notes}
- [45 seconds]
- [CLICK] The package auto-generates bias mitigations based on your specific dashboard context and user personas
- [CLICK] Look at the specific suggestions...anchoring, framing, confirmation bias
- These aren't abstract concepts...they're bugs in human cognition that affect every user
- [CLICK] Same dashboard, gain framing..."We're making progress!" Optimistic, forward-looking
- [CLICK] Loss framing..."We're behind target!" Creates urgency, drives action
- Same data...but different psychological impact
- Marketing sees growth opportunities...Finance sees risk exposure...both are right, both need different framing
- bidux helps you present both perspectives fairly...or choose the right frame for your audience
- It also suggests things like reference lines for anchoring...confidence intervals to combat overconfidence...and alternative views to fight confirmation bias
- This is proactive debugging for human psychology
- [end by 10:10]
:::

@@

::: {.notes}
- [65 seconds]
- At this point...we know what the dashboard is for, what problems to fix, and what biases to guard against
- Now it's time to STRUCTURE the information flow
- [CLICK] Look at these implementation suggestions...specific bslib and shiny components
- Not just "reduce complexity" but "use bslib::accordion for progressive disclosure" with actual code examples...(with more detailed suggestions coming soon!)
- The package knows which components map to which psychological principles
- [CLICK] See this F-pattern heatmap? This is how users scan dashboards in Western cultures
- Top-left gets 100% attention...bottom-right might as well not exist
- Structure stage puts your most important insights where eyes naturally go
- Critical KPIs or Summaries go on the top...supporting details follow the F-pattern...deep-dive options progressively disclosed
- This is cognitive engineering...you're structuring information to match how the human brain processes visual data
- Your most expensive computations should align with where users actually look
- [end by 11:15]
:::

@@

::: {.notes}
- [40 seconds]
- Finally...we reach the last stage: validate and empower users
- This helps you know the dashboard works as intended...and helps users know what to do next
- It guides you through testing changes and iterating...but also suggests ways to empower user engagement
- For example...executive summaries tailored to personas...detailed wording for analysts, high-level for execs
- This stage generates checklists...checks accessibility...and creates those "aha" moments
- It's the difference between "I looked at the dashboard" and "I know exactly what to do now"
- But Validation isn't the end...it feeds back to Interpret...creating continuous improvement that you can always revisit as needed
- Now let's take a look at the full transformation
- [end by 11:55]
:::

@@

::: {.notes}
- [25 seconds]
- Here's our dashboard disaster again...18 filters, 6 tabs, information overload incarnate
- Technically perfect...every edge case handled, every possible view available
- Psychologically broken...users spent more time figuring out the interface than finding insights
- This dashboard had everything except users
- Let's see what behavioral science can do...
- [end by 12:20]
:::

@@

:::: {.r-stack}

@@

::: {.text-box-overlay-dark .fragment}
| Metric                  | Before (%) | After (%) |   Δ    | Relative change |
| ----------------------- | :--------: | :-------: | :----: | :-------------- |
| Visited app URL         |     95     |     97    |   +2   | +2%             |
| Opened dashboard        |     88     |     95    |   +7   | +8%             |
| Applied ≥1 filter       |     52     |     72    |   +20  | +38%            |
| Returning within 1 week |     20     |     35    |   +15  | +75%            |

::: {.fragment .text-xl style="margin-top: 30px;"}
**More users. Deeper engagement. Stronger retention.**
:::
:::
::::

::: {.notes}
- [65 seconds]
- Same functionality...but a transformed user experience
- Clean executive summary at the top...users know immediately what matters and what to do about it
- Progressive disclosure...complexity is available but not required
- Smart defaults based on actual usage patterns...not assumptions
- [CLICK] Now look at these metrics...this is the ROI of behavioral science
- Filter usage up 38%...people are actually exploring data instead of abandoning ship
- Return rate up 75%...they're coming back!
- That's the difference between a proper tool and shelfware
- This isn't magic...or AI...it's engineering for how humans actually think
- We reduced cognitive load...aligned with mental models...and created clear information hierarchy
- [CLICK] More users...deeper engagement...stronger retention...because we removed obstacles instead of adding features
- And as a developer...this also means...
    - Less support tickets because users can figure it out themselves
    - Fewer feature requests because they're finding what they need
    - Better performance reviews because your dashboards driving value
- And that's the difference between building features and products people love
- [end by 13:25]
:::

@@

@@

metrics_long <- tibble(
  Metric = factor(
    c("Time to Insight", "Task Completion", "User Satisfaction", "Return Rate"),
    levels = c("Time to Insight", "Task Completion", "User Satisfaction", "Return Rate")
  ),
  Before = c(180, 48, 44, 22),
  After  = c(115, 62, 56, 30)
) |>
  pivot_longer(
    cols = c(Before, After),
    names_to = "State", values_to = "Value"
  ) |>
  mutate(State = factor(State, levels = c("Before", "After"))) |>
  group_by(Metric) |>
  mutate(
    BeforeValue = Value[State == "Before"][1],
    DiffPP      = ifelse(State == "After", Value - BeforeValue, NA_real_),
    MaxValue    = max(Value)
  ) |>
  ungroup()

@@

ggplot(metrics_long, aes(x = Metric, y = Value, fill = State)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.7) +
  geom_text(
    aes(
      label = ifelse(
        Metric == "Time to Insight",
        paste0(Value, "s"),
        paste0(Value, "%")
      )
    ),
    position = position_dodge(width = 0.7),
    vjust = -0.5,
    size = 6,
    fontface = "bold"
  ) +
  geom_text(
    data = metrics_long |>
      filter(State == "After" & !is.na(DiffPP)) |>
      distinct(Metric, .keep_all = TRUE),
    aes(
      y = MaxValue + 20,
      label = ifelse(
        Metric == "Time to Insight",
        paste0(round(DiffPP), "s"),
        paste0("+", round(DiffPP), "%")
      )
    ),
    color = "#FF4081",
    size = 6,
    fontface = "bold"
  ) +
  scale_fill_manual(values = pal) +
  labs(
    title = paste("BID Framework Impact:", n_deploy, "Enterprise Deployments"),
    subtitle = "Average across healthcare, finance, and customer goods sectors",
    y = NULL,
    x = NULL
  ) +
  theme_minimal(base_size = 18) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.title = element_text(size = 24, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.text = element_text(size = 14),
    panel.grid.major.x = element_blank()
  ) +
  coord_cartesian(
    ylim = c(0, max(metrics_long$MaxValue) + 20)
  )
```

::: {.notes}
- [45 seconds]
- And here are just some data to back that up
- These aren't cherry-picked results...they the 12 enterprise deployments across industries I've worked on in the last year using the BID framework
- 65 seconds faster to first insight...that's over a minute saved per user, per session
- Multiply that by hundreds of users, daily visits...you're saving hours of collective time
- Task completion up from 48% to 62%...nearly a third more decisions actually getting made
- User satisfaction up 27%...that's the difference between "I have to use this" and "I want to use this"
- Why does this work across industries?...Because human psychology doesn't change across domains
- Your finance users have the same 7±2 working memory slots as healthcare users
- They all suffer from choice overload...
- The framework adapts to your domain...but the principles remain constant
- Better psychology equals better engineering equals happier developers and users
- [end by 14:10]
:::

@@

summary(result)

@@

::: {.fragment}
Let's do that together!
:::
:::

>>>>

---
title: "Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science"
author: "Jeremy Winget, PhD"
date: "2025-09-17"
date-format: "MMMM D, YYYY"
execute:
  echo: true
  eval: true
format:
  revealjs:
    code-block-background: true
    highlight-style: github
    slide-number: false
    transition: fade
    auto-animate: true
    width: 1600
    height: 900
    footer: "posit::conf(2025) | jrwinget\\.com"
    title-slide-attributes:
      data-background-image: "img/posit-conf.png"
      data-background-position: "top left"
      data-background-repeat: "no-repeat"
      data-background-color: "#213d4f"
      data-background-opacity: "0.25"
      data-background-size: "45%"
      data-notes: |
        [10 seconds; smile] Good morning everyone! Let's jump right in...
---

@@

:::: {.columns .center}
::: {.column width="45%" .fragment .center}
<div class="tenor-gif-embed" data-postid="596341483704119392" data-share-method="host" data-aspect-ratio="1.32447" data-width="100%"><a href="https://tenor.com/view/monkey-computer-work-bug-developer-gif-596341483704119392">Monkey Computer GIF</a>from <a href="https://tenor.com/search/monkey-gifs">Monkey GIFs</a></div><script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::

::: {.column width="45%" .fragment .center}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.32447" data-width="100%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div><script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::
::::

::: {.notes}
- [30 seconds]
- Who here has opened a dashboard with 18 filters across 6 tabs...and wondered where to even start?...Or maybe what they were even looking at?
- I see some knowing looks
- Now I'd like you to imagine you're NOT a data scientist...You're an executive who needs to make a decision in the next 10 mins
- [CLICK] This might be you...frantically clicking, hoping something makes sense
- [CLICK] Or maybe you have some...bigger feelings
- The truth is...we've all probably felt this way towards a dashboard...and it's never a good feeling
- [end by 40 secs]
:::

@@

::: {.notes}
- [60 seconds]
- So...welcome to my dashboard support group!...Let's do a quick check-in
- Raise your hand if you've designed a UI with 10+ filters
    - Keep it up if you've gone over 20?
- How about this...who's had a user ask for "just one more dropdown" when you KNEW it was just too much?
- [CLICK] If you're nodding along...you're among friends...No judgment here
- [CLICK] ...I'm Jeremy...and I've been there too
- I'm a psychologist by training...but probably not the kind that first comes to mind
- See...a lot of folks think psychology is all about couches and therapy...but the discipline is powered by data
- [CLICK] My journey took me from studying how groups make decisions under pressure...to building the systems that caused that pressure...to finally figuring out how to fix them
- And after years of building dashboards that were technically great...but didn't seem to land with users...I had this realization
- [end by 1:40 mins]
:::

@@

::: {.notes}
- 20 seconds
- A lot of dashboard failures aren't technical...They're psychological
- They don't fail bc of the wrong chart type
- They fail because humans need to use them...and humans are complicated
- But...we can engineer for psychology just like we engineer for performance
- Let me show you what I mean
- [end by 2:00]
:::

## Real client. Real problems. {background-image="img/ui-before1.png" background-size="contain" background-color="#2c3e50"}

::: {.text-box-overlay-dark .absolute left="-200" bottom="100"}
- 18 filters across 6 tabs
- Every possible data view
- Curiosity quickly vanished. So did visits.

@@

::: {.notes}
- [45 seconds]
- This is a dashboard from a client of mine...details blinded of course
- They had built a really impressive dashboard...
- Lots of filters...customized views...and some really cool visualizations for deep dives
- The client loved it bc..."users could analyze anything they want!"
- Except...they never did
- The opened the app...but and shortly after...completely deserted it
- They were finding creative workarounds...like Excel
- [CLICK] So we decided to roll out a minor update...and collect anonymous usage data with telemetry for 30 days..to get a better idea of what users were actually doing
- [end by 2:45]
:::

@@

::: {.notes}
- [35 seconds]
- And...it painted this user journey cliff of despair
- 95% of users visited the URL...fantastic! Shows people are interested
    - 88% opened the dashboard...also good...they're getting past the landing page
- But here's where it gets brutal...
- Only half applied a single filter...Just one!
    - And only 1/5 returned within 7 days
- This chart is behavioral science screaming "STOP DOING THIS TO YOUR USERS!"
- But how do we know what needs fixing?
- [end by 3:20]
:::

@@

::: {.notes}
- [75 seconds]
- With behavioral science!...which is the science of how humans think, feel and act...often in terms of decision making
- And decades of research shows people are often irrational...but in consistent ways
- For us as developers...this means our strange user behaviors are predictable...and we can design for them
- Here's a classic study that illustrates what I mean
- [CLICK] Researchers set up a tasting booth at a grocery store...and they varied the number of jams they had on display
- [CLICK] In the big display condition...there were 24 jams on the table
    - And in the small display condition...there were 6 jams
- [CLICK] The big display with 24 jams attracted more shoppers than the small display with 6 jams
    - BUT...only 3% of folks who stopped at the larger display bought jam...whereas 30% of people bought jam in the small display condition
- [CLICK] That's a 10x conversion difference!
- So more choice attracted attention but paralyzed action...
    - And filter-heavy dashboards do the exact same thing
- [end by 4:30]
:::

@@

- **Cognitive Load**
    - Too many choices overwhelm users' mental capacity

- **Progressive Disclosure**
    - Reveal complexity gradually as users need it

- **Framing Effects**
    - Same data can tell completely different stories
:::

@@

- **Cognitive Load**
    - Start with 3 key filters, hide 15 advanced ones

- **Progressive Disclosure**
    - Show summary → detailed dive → comparisons

- **Framing Effects**
    - "Revenue up 15%" vs "Revenue missed by 5%"
:::
::::

::: {.notes}
- [75 seconds]
- Now...the paradox of choice is just one of many behavioral science concepts we can apply to dashboards
- But these aren't abstract theories...they're actually engineering principles we can leverage
- Cognitive Load...this means too many choices overwhelm users' mental capacity
    - Like RAM...your users' brains have processing limits...every filter costs processing power
- Progressive Disclosure...this means revealing complexity gradually as users need it
    - Show vital information first...and reveal more as needed
    - Like lazy loading for human cognition
- Framing Effects...this is how the same data can tell completely different stories
    - "Revenue up 15%" versus "Missed target by 5%"
    - Identical data...but the Proj Manager might see success while the CFO sees failure
- [CLICK] In practice...this means starting with 3 filters not 18...showing summaries before details...and using smart defaults
- This is how we engineer the user experience as carefully as we would the backend
- But what if you don't know behavioral science?
- [end by 5:35]
:::

@@

::: {.notes}
- [45 seconds]
- Well...I bet you do know something about road trips
- And engineering apps with behavioral science is a lot like planning a road trip
- You need to know your destination...anticipate the hazards...and make sure everyone gets there in one piece
- That's where the Behavioral Insight Design (BID) framework comes in...[CLICK]
- Five simple stages that guide you through the process...and help prevent those dashboard disasters we've all seen
- What makes the BID framework different from other UX advice is that it's a concrete...step-by-step framework built on decades of research
- Each stage feeds into the next...creating this continuous feedback loop that keeps improving your dashboard
- Plus...the whole thing is bundled into an R package that handles the heavy lifting for you
- [end by 6:20]
:::

@@

::: {.notes}
- [45 seconds]
- Meet bidux...designed specifically for our workflows as Shiny developers
- It works with any dashboard you've already built...no refactoring, no starting over
- If you have telemetry?...Great!
    - It'll analyze actual user behavior and give you data-driven insights
- Or...if you don't have telemetry data...you can use simple verbal descriptions during the process
- Whatever inputs you provide...it will use them to auto-suggest improvements based on real research
- [CLICK] Think of it as having a behavioral scientist right in your console...always available, never judges, and doesn't charge by the hour
- Now...let me show you what this process actually looks like
- [end by 7:05]
:::

@@

::: {.notes}
- [65 seconds]
- First...we start with the INTERPRET stage...the why behind the dashboard
- [CLICK] Because the central question drives the focus for your users
    - It's what they're coming to the dashboard for in the first place
- [CLICK] And the data story helps create narrative structure for this
- Humans are hardwired for stories...not pivot tables
- Hook, context, tension, resolution...
    - It's the structure folks are looking for with their key insights...so we build it in from the start
- [CLICK] And finally...user personas help us devs take the user perspective
- The Product Manager probably doesn't think in SQL queries...They think in business outcomes
- So the dashboard needs to bridge that gap
- [CLICK] And putting all of that together...the function creates a summary for you to review before moving on
- If it tells you anything is missing...it could also be a good time to check in with your users
- [end by 8:10]
:::

@@

::: {.notes}
- [75 seconds]
- Bc when we don't consider their perspective...we end up building friction points into our dashboards
- Going back to my client's dashboard...here's what the filter usage looked like
- 14 filters hardly ever used...and 4 virtually ignored
- This is technical debt visualized
- Each unused filter is maintenance overhead...and there are probably database queries running for nothing bc of those unused filters
- That's why the NOTICE stage focuses on finding these friction points systematically
- [CLICK] The package auto-suggests relevant behavioral theories with confidence scores
- Here...90% confidence Cognitive Load Theory is relevant...too many choices paralyzing users
- But it could suggest Choice Overload...Analysis Paralysis...or Decision Fatigue...depending on your context
- This transforms "it's confusing" into specific engineering problems...like "cognitive load exceeds working memory"
- The package also links each theory to specific Shiny components that address it
- So you're not guessing anymore...you're addressing identified psychological barriers with targeted code changes
- [end by 9:25]
:::

@@

::: {.notes}
- [45 seconds]
- [CLICK] The package also suggests bias mitigations based on the dashboard context and user personas
- [CLICK] Look at theses suggestions...anchoring, framing, confirmation bias
- These are the bugs in human cognition that affect every user...but now we can guard for them
- In my client's case...they had multiple stakeholders with different perspectives
    - So we focused a lot on framing
- [CLICK] The progress (gain) frame shows the *relative difference* between the raw value and the benchmark...which highlighted the achievements the marketing team was interested in
- [CLICK] BUT...the gap (loss) framing shows the raw values vs. the benchmark...which focused on the improvements still needed that the product teams wanted
- [CLICK] Same data...but different psychological impact
- This is proactive debugging for human psychology
- [end by 10:10]
:::

@@

::: {.notes}
- [65 seconds]
- Okay...at this point...we know what the dashboard is for...the problems to fix...and what biases to guard against
- Now it's time to STRUCTURE the information flow
- [CLICK] And using bid_structure we get these specific implementation suggestions...
- Not just "reduce complexity" but "use bslib::accordion for progressive disclosure"
- The package knows which components map to which psychological principles...and gives multiple options to try
- For example...
- [CLICK] See this F-pattern heatmap? This is how most users scan dashboards in Western cultures
- Top-left gets 100% attention...bottom-right might as well not exist
- The Structure stage puts your most important insights where eyes naturally go
- So critical KPIs or summaries are placed at the top...supporting details follow the F-pattern...deep-dive options progressively disclosed
- This is cognitive engineering...we're structuring information to match how the human brain processes visual data
- [end by 11:15]
:::

@@

::: {.notes}
- [40 seconds]
- Finally...we reach the last stage...validate and empower users
- This helps you know the dashboard works as intended...and helps users know what to do next
- It guides devs through testing changes and iterating...but also suggests ways to empower user engagement
- For example...executive summaries tailored to personas...
    - Maybe we use more technical wording for analysts...but high-level for execs
- This stage also generates checklists...checks accessibility...and creates those "aha" moments
- It's the difference between "I looked at the dashboard"...and "I know exactly what to do now"
- But Validation isn't the end...it feeds back to Interpret...creating continuous improvement that you can always revisit as needed
- [PAUSE]
- Now let's take a look at the full transformation
- [end by 11:55]
:::

@@

::: {.notes}
- [25 seconds]
- Here's our dashboard disaster again...18 filters, 6 tabs, information overload incarnate
- Technically perfect...but psychologically broken
- The dashboard that had everything...except users
- Let's see what behavioral science can do to turn this around...
- [end by 12:20]
:::

@@

@@

::: {.text-box-overlay-dark .fragment .absolute left="288" bottom="200"}
| Metric                  | Before (%) | After (%) |   Δ    | Relative change |
| ----------------------- | :--------: | :-------: | :----: | :-------------: |
| Visited app URL         |     95     |     97    |   +2   | +2%             |
| Opened dashboard        |     88     |     95    |   +7   | +8%             |
| Applied ≥1 filter       |     52     |     72    |   +20  | +38%            |
| Returning within 1 week |     20     |     35    |   +15  | +75%            |

::: {.fragment .text-xl style="margin-top: 30px;"}
**More users. Deeper engagement. Stronger retention.**
:::
:::

::: {.notes}
- [65 seconds]
- Here's the same dashboard...but with a transformed user experience
- Clean executive summary at the top...users know immediately what matters and what to do about it
- Progressive disclosure...complexity is available but not required
- Smart defaults based on actual usage patterns...not assumptions
- [CLICK] And here are the data
- Filter usage up 38%...people are actually exploring data instead of abandoning ship
- Return rate up 75%...they're coming back!
- This is the ROI of behavioral science
- It's not magic...or AI...it's engineering for how humans actually think
- We reduced cognitive load...aligned with mental models...and created clear information hierarchy
- [CLICK] More users...deeper engagement...stronger retention...because we removed obstacles instead of adding features
- And as a developer...this also means...
    - Less support tickets because users can figure it out themselves
    - Fewer feature requests because they're finding what they need
    - Better performance reviews because your dashboards driving value
- And that's the difference between building features and products people love
- [end by 13:25]
:::

@@

library(patchwork)

@@

# summary metrics
metrics_long <- tibble(
  Metric = factor(
    c("Avg Session Duration", "Dashboard Load Time", "Daily Active Users", "Query Success Rate"),
    levels = c("Avg Session Duration", "Dashboard Load Time", "Daily Active Users", "Query Success Rate")
  ),
  Before = c(8.2, 4.8, 340, 76),  # minutes, seconds, users, percentage
  After  = c(10.8, 3.4, 425, 87)
) |>
  pivot_longer(
    cols = c(Before, After),
    names_to = "State", values_to = "Value"
  ) |>
  mutate(State = factor(State, levels = c("Before", "After"))) |>
  group_by(Metric) |>
  mutate(
    BeforeValue = Value[State == "Before"][1],
    DiffPP      = ifelse(State == "After", Value - BeforeValue, NA_real_),
    MaxValue    = max(Value)
  ) |>
  ungroup()

@@

# plot function
create_metric_plot <- function(metric_name) {
  data_subset <- metrics_long |> filter(Metric == metric_name)

  ggplot(data_subset, aes(x = Metric, y = Value, fill = State)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.7) +
    geom_text(
      aes(
        label = case_when(
          Metric == "Avg Session Duration" ~ paste0(Value, " min"),
          Metric == "Dashboard Load Time" ~ paste0(Value, "s"),
          Metric == "Daily Active Users" ~ as.character(Value),
          Metric == "Query Success Rate" ~ paste0(Value, "%")
        )
      ),
      position = position_dodge(width = 0.7),
      vjust = -0.5,
      size = 6,
      fontface = "bold"
    ) +
    geom_text(
      data = data_subset |>
        filter(State == "After" & !is.na(DiffPP)) |>
        distinct(Metric, .keep_all = TRUE),
      aes(
        y = MaxValue + (MaxValue * 0.15),
        label = case_when(
          Metric == "Avg Session Duration" ~ paste0("+", round(DiffPP, 1), " min"),
          Metric == "Dashboard Load Time" ~ paste0(round(DiffPP, 1), "s"),
          Metric == "Daily Active Users" ~ paste0("+", round(DiffPP)),
          Metric == "Query Success Rate" ~ paste0("+", round(DiffPP), "%")
        )
      ),
      color = "#FF4081",
      size = 6,
      fontface = "bold"
    ) +
    scale_fill_manual(values = pal) +
    labs(y = NULL, x = NULL) +
    theme_minimal(base_size = 18) +
    theme(
      legend.position = "top",
      legend.title = element_blank(),
      axis.text = element_text(size = 14),
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(size = 16, face = "bold")
    ) +
    coord_cartesian(
      ylim = c(0, max(data_subset$MaxValue) * 1.25)
    )
}

# individual plots
p1 <- create_metric_plot("Avg Session Duration")
p2 <- create_metric_plot("Dashboard Load Time")
p3 <- create_metric_plot("Daily Active Users")
p4 <- create_metric_plot("Query Success Rate")

# combine plots with shared legend
combined_plot <- (p1 + p2 + p3 + p4) +
  plot_layout(guides = "collect", ncol = 4) &
  theme(legend.position = "top")

combined_plot +
  plot_annotation(
    title = paste("Dashboard Analytics Impact:", n_deploy, "Enterprise Deployments"),
    subtitle = "Performance metrics across healthcare, finance, and consumer goods sectors",
    theme = theme(
      plot.title = element_text(size = 24, face = "bold"),
      plot.subtitle = element_text(size = 14)
    )
  )
```

::: {.notes}
- [45 seconds]
- And here are the results from 12 deployments I've worked on in the last year using the BID framework
- On average, session duration increased on by 2.6 minutes...which is the difference between "I have to use this" and "I want to use this"
- Faster load times...
- Multiply that by hundreds of users, daily visits...that's hours of collective time saved
- Daily active users increased by 85...
- Query success rate went up from 76% to 87%...
- And the reason this works is because human psychology is predictable
- Your finance users have the same working memory slots as your healthcare users
- The framework adapts to your domain...but the principles remain constant
- Engineering for psychology = better dashboard engineering = happier developers and users
- [end by 14:10]
:::

@@

@@

::: {.fragment}
Let's fix that together!
:::
:::
